Do researchers preferentially collaborate with same-gendered colleagues?

### 2018-05-23 manuscript/Manuscript_Claire.Rmd
manuscript/

---
title: "Do researchers preferentially collaborate with same-gendered colleagues?"
author:
- Luke Holman* and Claire Morandin
- '*\textit{luke.holman@unimelb.edu.au} \vspace{5mm}'
output:
  pdf_document: default
  html_document: default
bibliography: references.bib
fontsize: 12pt
header-includes: 
  \usepackage[all]{hypcap} 
  \usepackage{amsmath} 
  \usepackage{booktabs}
  \usepackage{caption} 
  \usepackage[labelfont=bf]{caption} 
  \usepackage{titling} 
  \pretitle{\begin{flushleft}\LARGE}
  \posttitle{\par\end{flushleft}\vskip 0.5em} 
  \preauthor{\begin{flushleft}\large \lineskip 0.5em} 
  \postauthor{\par\end{flushleft}} 
  \predate{\begin{flushleft}\large} 
  \postdate{\par\end{flushleft}}
  \usepackage{fancyhdr} 
  \pagestyle{fancy} 
  \fancyhead[LO,LE]{\textsl{\leftmark}} 
  \rhead[]{Conference gender ratios} 
  \usepackage[noabbrev,capitalise]{cleveref}
  \usepackage{titlefoot} 
  \usepackage{amssymb} 
  \usepackage{rotating}
csl: the-american-naturalist.csl
abstract: Evidence suggests that women in academia are hindered by conscious and unconscious biases, and that many female researchers feel excluded from formal and informal opportunities for research collaboration. In addition to ensuring fairness and helping redress gender imbalance in the academic workforce, increasing women???s access to collaboration could help scientific progress by drawing on more of the available human capital. Here, we test whether researchers preferentially collaborate with same-gendered colleagues, using more stringent methods and a larger dataset than in past work. Our results reaffirm that researchers co-publish with colleagues of the same gender, and show that this ???gender homophily??? is slightly stronger today than it was 10 years ago. Contrary to our expectations, we found no evidence that homophily is driven mostly by senior academics, and no evidence that homophily is strongest in fields where women are in the minority. Interestingly, homophily was negatively correlated with journal impact factor (standardised by research discipline), as predicted if mixed-gender teams produce better research. \vspace{5mm} \par\noindent \textbf{Keywords:} Gender bias, Homophily, Scientific collaboration, Text mining, Women in STEM.
---

@@

\maketitle\unmarkedfntext{*School of BioSciences, The University of Melbourne, Victoria, Australia}
\newpage

@@

# Discussion
We found evidence that researchers tend to publish with same-gendered coauthors, despite implementing stringent controls for ???spurious??? homophily resulting from Wahlund effects (\autoref{wahlund_plot}). Our data therefore reaffirm earlier conclusions [@Ferber_1980; @McDowell_1992; @Crow_2015; @Ghiasi_2015; @Araujo_2017a; @Fahmy_2017; @Jadidi_2017; @Zettler_2017; @Teele_2017] that research collaborations occur between same-gendered colleagues substantially more often than expected if collaborators are selected without regard to their gender. Many journals had $\alpha'$ values over 0.1, meaning that the gender ratio of the average man's coauthors is skewed towards men by >10%, relative to the gender ratio of a woman's coauthors. Few journals had $\alpha'$ values below zero, and fewer still had significantly negative $\alpha'$ values, suggesting that it is comparatively rare for researchers to preferentially select opposite-gendered collaborators.

We hypothesized that disciplines with strong gender-biases (e.g. physics and computer science for men and nursing, midwifery for women [@Holman_2018]) would display a higher level of gender-homophily due to the increased difficulty to find an opposite-gender co-author. However, we found no evidence that gender homophily is restricted to particular disciplines: instead, there was evidence of homophily across many diverse research disciplines (FIG). Gender homophily is thus a global concern rather than an artifact of the existing gender gap (REF). Additionally, we found that gender homophily was slightly stronger in research disciplines with a female-biased gender ratio, as compared to male-biased or gender-balanced fields (whether it is in the first or last author position). One interpretation of this result is that men are more likely to seek out other men in fields in which men are in the minority, relative to the homophily shown by women in fields in which women are in the minority. This male-driven homophily could be, in part, due to the shortage of opportunities for women to participate in international research projects (@Elsevier report),  or international conferences (REF) and therefore less likely to collaborate across gender or fields.

We also found that gender homophily was marginally stronger in 2015-2016, relative to 2005-2006. This result suggests that recent increases in the number of women working in STEMM, as well as recent efforts to recruit, retain, and celebrate women in STEMM, have co-occurred with rise in the frequency of same-gender collaborations, not a fall. We speculate that this trend might result, at least in part, from the increasing number of women working in senior positions in STEMM over the last 10 years [@Bendels_2018; @Long_2015; @McKenzie_2017]. Coupled with an ever increase proportion of women in the first author position [@West_2013], this trend leads to an increase in same-sex collaboration. Because many co-authorships involve a junior and a senior researcher (e.g. a student and professor), and the gender ratio is typically more skewed among senior staff [@Shaw_2012, @West_2013], there will be more opposite-gender collaborations than expected if the gender ratio were the same at both career stages. Thus, the observed increase in homophily might partially be due to the greater numbers of women working in leading roles. However, gender homophily during collaboration and co-authoring of scientific manuscript may also impact negatively women careers as there is still a smaller proportion of women in science and thus, making it harder to find female collaborators (Bentley and Adamson 2003).

We also found that journals with a high impact factor (relative to other journals in the same research discipline) tended to have weaker gender homophily than did low-impact journals. If one accepts the contentious claim that papers published in high-impact journals tend to be of higher quality than those in low-impact journals [e.g. @Garfield_2006], then this result provides correlational support for the hypothesis that mixed-gender teams produce better research than single-gender teams [@Campbell_2013]. However, journal impact factors tend to be negatively associated with the proportion of women [@Holman_2018. REF]. Contrary to our expectations, we found no evidence that homophily is driven mostly by senior academics, and no evidence that homophily is strongest in fields where women are in the minority. Interestingly, homophily was negatively correlated with journal impact factor (standardised by research discipline), as predicted if mixed-gender teams produce better research.

Academic publishing is one of the most (if not the most) important tool for a researcher to display their work, and is frequently used as a success indicator for being promoted, hired as a faculty member, or being invited as plenary speaker. Our finding that researchers are more likely to publish with same-gendered coauthors implies that co-authorship listing is not a random process, but instead creates additional gender imbalance. Homophily could make it harder for women to find collaborators in disciplines where women are in the minority [i.e. almost all STEMM fields; @Holman_2018]. Because women make up for as little as 30% of the STEMM scientist community [@Holman_2018], they are therefore less likely to be picked as coauthors, and, as a result, gender homophily creates further gender imbalance and disadvantages for women. Men tend to publish more in high impact journals as first author [@West_2013], and generate more citations than papers lead by first author women [@Long_1992; @Lariviere_2013]. And as a consequence, gender homophily is also likely to contribute to lower publication rates for women [@McDowell_1992], and lower scientific success (REF).

Although the proportion of women in science has significantly increased over the last decades (REF), women still tend to leave their academic careers behind considerably more than their male counterpart (D??barre et al 2018). Lack of scientific recognition (Lincolm et al- 2012), research funding (Bozeman and Corley 2004), travel opportunities, invitation to speak at scientific events (Isbell et al. 2012; Schroeder et al- 2013; Klein et al- 2017), discrimination during hiring (Moss-Racusin et al- 2012; Reuben et al. 2014), when writing a recommendation letter (Trix and Psenka 2003; Schmader et al- 2007), or personal reasons (e.g. child rearing or family ties (Zimmer et al- 2006)), all in all this concentration of small disadvantages drives women out of science. Whether their decision to leave academia is deliberate or not, it will influence the future of the scientific community, as mentoring, networking, and supporting younger women scientists, are especially important for women in academia.

Why does gender homophily exist in science? Collaboration often leads to co-authorship publications, citations and additional professional recognitions, therefore first author should always be considering collaborators that are most likely to enhance their productivity and in the long term also their academic success, rather than pick co-authors according to their gender. First of all, it is likely that decades ago, the probability to meet a woman colleague working on the same area of expertise was slimmer than now. However, the number of women in STEM is increasing, and it is not likely to be an obstacle to collaboration nowadays.

>>>>

---
title: "Do researchers preferentially collaborate with same-gendered colleagues?"
author:
- Luke Holman* and Claire Morandin+
- '*\textit{luke.holman@unimelb.edu.au} \vspace{5mm}'
output:
  pdf_document: default
  html_document: default
bibliography: references.bib
fontsize: 12pt
header-includes: 
  \usepackage[all]{hypcap} 
  \usepackage{amsmath} 
  \usepackage{booktabs}
  \usepackage{caption} 
  \usepackage[labelfont=bf]{caption} 
  \usepackage{titling} 
  \pretitle{\begin{flushleft}\LARGE}
  \posttitle{\par\end{flushleft}\vskip 0.5em} 
  \preauthor{\begin{flushleft}\large \lineskip 0.5em} 
  \postauthor{\par\end{flushleft}} 
  \predate{\begin{flushleft}\large} 
  \postdate{\par\end{flushleft}}
  \usepackage{fancyhdr} 
  \pagestyle{fancy} 
  \fancyhead[LO,LE]{\textsl{\leftmark}} 
  \rhead[]{Conference gender ratios} 
  \usepackage[noabbrev,capitalise]{cleveref}
  \usepackage{titlefoot} 
  \usepackage{amssymb} 
  \usepackage{rotating}
csl: the-american-naturalist.csl
abstract: Evidence suggests that women in academia are hindered by conscious and unconscious biases, and that many female researchers feel excluded from formal and informal opportunities for research collaboration. In addition to ensuring fairness and helping redress gender imbalance in the academic workforce, increasing women???s access to collaboration could help scientific progress by drawing on more of the available human capital. Here, we test whether researchers preferentially collaborate with same-gendered colleagues, using more stringent methods and a larger dataset than in past work. Our results reaffirm that researchers co-publish with colleagues of the same gender, and show that this ???gender homophily??? is slightly stronger today than it was 10 years ago. Contrary to our expectations, we found no evidence that homophily is driven mostly by senior academics, and no evidence that homophily is strongest in fields where women are in the minority. Interestingly, homophily was negatively correlated with journal impact factor (standardised by research discipline), as predicted if mixed-gender teams produce better research. \vspace{5mm} \par\noindent \textbf{Keywords:} Gender bias, Homophily, Scientific collaboration, Text mining, Women in STEM.
---

@@

\maketitle\unmarkedfntext{*School of BioSciences, The University of Melbourne, Victoria, Australia}
\maketitle\unmarkedfntext{+Organismal and Evolutionary Biology Research Programme,
Faculty of Biological and Environmental Sciences, University of Helsinki, Finland}
\newpage

@@

# Discussion
We found evidence that researchers tend to publish more with same-gendered coauthors, despite implementing stringent controls for ???spurious??? homophily resulting from Wahlund effects (\autoref{wahlund_plot}). Our data therefore reaffirm earlier conclusions [@Ferber_1980; @McDowell_1992; @Bentley_2003; @Crow_2015; @Ghiasi_2015; @Araujo_2017a; @Fahmy_2017; @Jadidi_2017; @Teele_2017; @Zettler_2017] that research collaborations occur between same-gendered colleagues substantially more often than expected if collaborators are selected without regard to their gender. Many journals had $\alpha'$ values over 0.1, meaning that the gender ratio of the average man's coauthors is skewed towards men by >10%, relative to the gender ratio of a woman's coauthors. Few journals had $\alpha'$ values below zero, and fewer still had significantly negative $\alpha'$ values, suggesting that it is comparatively rare for researchers to preferentially select opposite-gendered collaborators.

As some disciplines exhibit strong gender-biases (e.g. physics and computer science with predominantly men co-authors, and nursing, midwifery for women [@Holman_2018]), we hypothesized that these disciplines would also display a higher level of gender-homophily due to the increased difficulty to find an opposite-gender co-author. However, we found no evidence that gender homophily is restricted to particular disciplines: instead, there was evidence of homophily across many diverse research disciplines (Figure 2). This finding may be indicative that gender homophily in science is a comprehensive concern, rather than an artifact of the existing gender gap, and not limited to stereotypical disciplines. Furthermore, we found that gender homophily was slightly stronger in research disciplines with a female-biased gender ratio, as compared to male-biased or gender-balanced fields (whether it is in the first or last author position). One interpretation of this result is that men are more likely to seek out other men in fields in which men are in the minority, relative to the homophily shown by women in fields in which women are in the minority.

We also found that gender homophily was marginally stronger in 2015-2016, relative to 2005-2006. This result suggests that recent increases in the number of women working in STEMM, as well as recent efforts to recruit, retain, and celebrate women in STEMM, have co-occurred with rise in the frequency of same-gender collaborations, not a fall. We speculate that this trend might result, at least in part, from the increasing number of women working in senior positions in STEMM over the last 10 years [@Bendels_2018; @Long_2015; @McKenzie_2017]. Because many co-authorships involve a junior and a senior researcher (e.g. a student and professor), and the gender ratio is typically more skewed among senior staff [@Shaw_2012, @West_2013], there will be more opposite-gender collaborations than expected if the gender ratio were the same at both career stages. Thus, the observed increase in homophily might partially be due to the greater numbers of women working in leading roles.

Despite the fact that journal impact factors tend to be negatively associated with the proportion of women [@Holman_2018], we found that journals with a high impact factor (relative to other journals in the same research discipline) tended to have weaker gender homophily than did low-impact journals (Figure 4). If one accepts the contentious claim that papers published in high-impact journals tend to be of higher quality than those in low-impact journals [e.g. @Garfield_2006], then this result provides correlational support for the hypothesis that mixed-gender teams produce better research than single-gender teams [@Campbell_2013].

Why does gender homophily exist in science? Collaboration often leads to co-authorship publications, citations and additional professional recognitions, therefore first author should always be considering collaborators that are most likely to enhance their productivity and in the long term also their academic success, rather than pick co-authors according to their gender. First of all, it is likely that decades ago, the probability to meet a woman colleague working on the same area of expertise was slimmer than now. Even though the number of women in STEM is increasing, and it is not likely to be an obstacle to collaboration nowadays; we cannot forgo that female representation in science is still not equal to that of males. Second, the most basic source of homophily is space, scientists are more likely to collaborate with colleagues closer to them in geographic location (@McPherson_2011). In the case of women, personal considerations (e.g. family commitments) represent the principal restrictions on their international mobility. As a results, women are less likely to originate collaborations with foreign colleagues [@Lemoine_1992; @Lewison_2001; @ Webster_2001; @ Lariviere_2011]. Lastly, men and women think differently on average (REF), and are more likely to initiate collaborations based on similarity of thoughts to progress their scientific output (REF).

Academic publishing is one of the most (if not the most) important tool for a researcher to display their work, and is frequently used as a success indicator for being promoted and hired as a faculty member. Our finding that researchers are more likely to publish with same-gendered coauthors implies that co-authorship listing is not a random process, but instead creates additional gender imbalance. Homophily could make it harder for women to find collaborators in disciplines where women are in the minority [i.e. almost all STEMM fields; @Holman_2018]. Because women make up for as little as 30% of the STEMM scientist community [@Holman_2018], they are therefore less likely to be picked as coauthors, and, as a result, gender homophily creates further gender imbalance and disadvantages for women. Men tend to publish more in high impact journals as first author [@West_2013], and generate more citations than papers lead by first author women [@Long_1992; @Lariviere_2013]. And as a consequence, gender homophily is also likely to contribute to lower publication rates for women [@McDowell_1992].

Although the proportion of women in science has significantly increased over the last decades, women still tend to leave their academic careers behind considerably more than their male counterpart [@Debarre_2018] . Lack of scientific recognition [@Lincoln_2012], lack of research funding [@Bozeman_2004], lack of invitation to speak at scientific events[@Isbell_2012; @Schroeder_2013; @Klein_2017], discrimination during hiring [@MossRacusin_2017; @Reuben_2014], discrimination when writing a recommendation letter [@Trix_2003; @Schmader_2007], all in all this concentration of small disadvantages drives women out of science. Whether their decision to leave academia is deliberate or not, it will negatively impact science overall, as mentoring, networking, and supporting younger women scientists are especially important for women in academia.

Altogether, our findings suggest that researchers preferably collaborate with same-gendered colleagues in STEMM research. Despites using a more stringent analysis methods (e.g. Wahlund effects) and the largest dataset to date, our results were in accord with longstanding and recent research. Gender homophily in science not only reinforces the less favorable status of women in science, but also creates long term problems (see above). It is important to explore the causes and implement measures to redress the gender disparity. We advocate for the introduction and implementation of gender policies in order to support opposite sex collaboration, to give men and women a chance to collaborate across gender, fields, and to promote scientific and innovative excellence.
